---
title: "Conditional-and-Interventional-Distribution"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Conditional-and-Interventional-Distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bnSEM)
```

Bayesian Networks allow investigating the conditional and interventional 
distribution of network nodes. 

The central difference between conditional and interventional distribution
can be explained with a variation of the classical sprinkler-rain model [see e.g.,
@murphy2012machine, p. 944]. Assume that R denotes rain, W is the weather report,
and S is the activation of water sprinklers. In the model, both the weather report
and the activation of sprinklers are affected by rain:
```{r, echo=FALSE}
bnlearn::model2network("[R][W|R][S|R]") |>
  plot()
```


- The conditional distribution $P(W = \text{rain} |S  = \text{activated})$ will tell us the probability 
of the weather report reporting rain on a day where we observe that the sprinklers are activated. 
Because sprinklers will typically not be activated on days where it rains, we will know that
the weather forcast will probably also report rain.
- The interventional distribution $P(W | do(s))$ will tell us the probability 
of the weather report reporting rain when we activate (i.e., intervene on) the
sprinklers. Of course, activating or deactivating the sprinkler will not have any
effect on the weather report because the connections between both is caused by a 
third variable.

The difference between conditional and interventional distribution in SEM is discussed in
detail by @gische2022beyond.

## Model

We will use a cross-lagged panel model for our example. We will assume that the
correct model is given by



# Bibliography
